---
title: "Cluster correspondence across repeated maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cluster correspondence across repeated maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Cluster labels are arbitrary identifiers. A cluster called `A` at one period
may be called `Y` in the next even if its membership has not changed.
`match_clusters()` assigns persistent bookkeeping identifiers using shared
entity membership in adjacent periods. It accepts already supplied labels;
it does not fit a clustering algorithm or infer that a cluster is a stable
population. Coordinate alignment and cluster correspondence solve different
problems, and matching does not require an aligned object.

## Match supplied labels

```{r example}
library(driftmapR)

first <- data.frame(
  entity = letters[1:8], time = 1,
  x = c(0, 1, 0, 1, 4, 5, 4, 5), y = rep(c(0, 0, 1, 1), 2),
  cluster = rep(c("A", "B"), each = 4)
)
second <- transform(first, time = 2, cluster = rep(c("Y", "X"), each = 4))
fit <- drift_data(rbind(first, second)) |> match_clusters()
fit$clusters
fit$diagnostics$cluster_matching$assignments
fit$diagnostics$cluster_matching$periods

# Labels change, but all eight entities retain their persistent identifiers.
members <- fit$clusters
before <- members[members$period_index == 1, ]
after <- members[members$period_index == 2, ]
stopifnot(identical(before$cluster_id,
                    after$cluster_id[match(before$entity, after$entity)]))
```

`cluster_original` preserves the supplied label, while `cluster_id` is the
persistent identifier. `cluster_status` distinguishes the first period,
inherited identifiers, newly allocated identifiers, and unassigned rows.
Use the original labels for provenance and the persistent identifiers for
descriptive longitudinal comparisons.

## The matching objective and denominator

At an adjacent boundary, let $S$ be entities observed and assigned a nonmissing,
non-noise label in **both** periods. For predecessor cluster $a$ and current
cluster $b$, define

$$
n_{ab}=\#\{i\in S:c_{i,t-1}=a,\ c_{it}=b\},
\qquad
J_{ab}=\frac{n_{ab}}{n_{a\cdot}+n_{\cdot b}-n_{ab}}.
$$

The two marginal counts use the same jointly labeled shared cohort $S$.
Entrants, exits, and an entity unassigned at either endpoint do not contribute
to these overlap scores. Therefore this Jaccard score describes agreement in
the available overlap cohort, rather than full-population membership stability.
The period diagnostics expose the available denominator.

After removing zero-overlap edges and edges below `min_overlap` or
`min_jaccard`, the assignment maximizes

$$
\sum_{a,b} n_{ab}m_{ab},\qquad
m_{ab}\in\{0,1\},\quad
\sum_b m_{ab}\leq 1,\quad \sum_a m_{ab}\leq 1.
$$

Clusters may remain unmatched. Jaccard is an eligibility threshold; the
optimization objective is the **total shared-entity count**, not the sum of
Jaccard scores. This is a global one-to-one optimization, so the largest
individual overlap need not be selected if competing edges yield a larger
total. Thresholds are applied before optimization.

```{r stricter}
strict <- drift_data(rbind(first, second)) |>
  match_clusters(min_overlap = 3L, min_jaccard = 0.75)
strict$diagnostics$cluster_matching$transitions
```

## Ambiguity and diagnostics

The diagnostics retain assignments, all positive membership flows, the cluster
registry, period summaries, and overlap/score matrices. A chosen edge is a
bookkeeping decision. It is not a significance test or a probability that two
clusters are the same.

When equal-scoring assignments exist, the exact tie rule chooses the
lexicographically earliest sequence of current original labels in predecessor
persistent-ID order, with unmatched choices last. This rule is reproducible but
the spelling of tied labels can change which equally good correspondence is
chosen. An assignment margin of zero indicates that another optimum exists
without the chosen edge; it is not a confidence measure. Disabling tie
diagnostics avoids the extra diagnostic solves, but does not change the
objective or deterministic tie rule.

```{r ambiguity}
tie_first <- first[1:4, ]
tie_first$cluster <- c("A", "A", "B", "B")
tie_second <- transform(tie_first, time = 2, cluster = c("X", "Y", "X", "Y"))
tied <- drift_data(rbind(tie_first, tie_second)) |> match_clusters()
tied$diagnostics$cluster_matching$assignments
tied$diagnostics$cluster_matching$periods
```

## Splits, merges, missing labels, and turnover

One-to-one assignment cannot transmit one predecessor identifier to both
children of a split, or combine two identifiers into one. One child can inherit
the old identifier; unmatched children receive new ones. At a merge, one
predecessor can retain its identifier. Inspect the full transition table to
avoid mistaking the selected correspondence for a complete account of change.

Multiple positive outgoing overlaps flag a `split_candidate`; multiple positive
incoming overlaps flag a `merge_candidate`. These flags describe observed
overlap structure, including edges excluded by thresholds. Entities switching
between otherwise continuing clusters can create both flags. Neither flag
establishes a generating split or merge process.

```{r split}
split_first <- transform(first, cluster = "A")
split_second <- transform(first, time = 2, cluster = c(rep("X", 6), rep("Y", 2)))
split_fit <- drift_data(rbind(split_first, split_second)) |> match_clusters()
split_fit$diagnostics$cluster_matching$transitions
```

Missing labels are unassigned. A supplied noise label is treated as unassigned
only when explicitly included in `noise`, for example `noise = "-1"`.
Unassigned rows retain no persistent identifier and do not count toward the
matching denominator. By default, any valid nonmissing label, including `"0"`,
is an ordinary cluster label.

An unmatched current cluster receives a fresh identifier even if its original
label appeared earlier. Matching uses only adjacent periods: a lost identifier
is not resurrected after a gap. The same raw label in two periods does not
override missing or inadequate shared membership. A `new` status means unmatched
under the current rules; it does not prove that a new underlying population
has formed.

## Reproduce the synthetic validation

The source checkout includes `data-raw/simulate-clusters.R`. After installing
the package, run from the checkout:

```{bash, eval=FALSE}
Rscript data-raw/simulate-clusters.R ../cluster-simulation-output
```

Eight small fixtures cover permutations, member switching, a split, a merge,
entrants/exits, exact ties, disjoint entity sets, and a conflict where a greedy
match fails. A five-period panel adds sequential combinations, with 96
observations per period and 136 unique entities. Independently specified parent
correspondences check identifier preservation. Exhaustive enumeration on these
small overlap matrices checks the maximum objective and whether alternative
optimal assignments exist. The script stops if any check disagrees and saves
inspectable inputs, truth, outputs, diagnostics, matrices, and figures.

These are controlled descriptive checks, not a bootstrap stability study.
Cluster fitting, probabilistic split/merge inference, bootstrap cluster
stability, and movement uncertainty remain separate development targets.
