---
title: "PCA and classical-MDS adapters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA and classical-MDS adapters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`embed_snapshots()` independently fits a two-dimensional embedding at each
period and returns an **unaligned** `driftmap`. PCA uses the singular value
decomposition. Classical multidimensional scaling (classical MDS, `"cmds"`)
uses the eigenvectors of the centered squared-distance matrix. Both retain the
original input, fitting metadata and diagnostics for later inspection.
Alignment remains an explicit subsequent operation.

## Start from repeated feature measurements

Select the feature columns explicitly. Metadata such as cluster labels should
not accidentally become measured dimensions. Entity identities and period
ordering follow the same rules as `drift_data()`. Features must be finite
numeric columns observed in every row; no imputation is performed.

```{r feature-data}
library(driftmapR)
first <- data.frame(entity = letters[1:8], time = 1,
                    u = c(-2, -1, 0, 1, 2, 3, 1, -1),
                    v = c(0, 1, -1, 2, 0, 1, 3, -2))
second <- transform(first, time = 2)
second$u[second$entity == "h"] <- second$u[second$entity == "h"] + 0.6
second$v[second$entity == "h"] <- second$v[second$entity == "h"] - 0.8
panel <- rbind(first, second)
# An isometric loading into four dimensions makes the movement truth exact.
panel$f1 <- panel$u / sqrt(2)
panel$f2 <- panel$v / sqrt(2)
panel$f3 <- panel$u / sqrt(2)
panel$f4 <- panel$v / sqrt(2)
features <- paste0("f", 1:4)
input <- panel[c("entity", "time", features)]
raw <- embed_snapshots(input, features = features, method = "pca")
raw$diagnostics$embedding
```

The embedding axes may change sign or rotate between independent fits. Changes
in the retained subspace can also occur when a higher-dimensional geometry is
truncated to two dimensions. Procrustes alignment removes rigid coordinate
indeterminacy; it cannot restore geometry discarded by the embedding.

```{r align}
# These seven entities are known to be stable in this constructed example.
fit <- align_snapshots(raw, anchors = letters[1:7], scale = FALSE)
movement <- measure_drift(fit)
movement[c("entity", "time_from", "time_to", "distance")]
stopifnot(max(movement$distance[movement$entity != "h"]) < 1e-8,
          abs(movement$distance[movement$entity == "h"] - 1) < 1e-8)
```

Entity `h` moves by $(0.6,-0.8)$, a Euclidean distance of one. All other entities
remain stable. Anchors encode that known stability; using moving entities in
the alignment fit can absorb part of their movement. Centering and alignment
also remove shared translation, so a global feature-level shift is not an
identifiable entity drift effect in this workflow.

```{r plot, fig.width=6, fig.height=4}
plot_drift_map(fit)
```

## Preprocessing defines the distance

Every snapshot is centered. `standardize` chooses the feature standard
deviations used to divide the data before fitting:

- `"none"` preserves supplied feature units and is the default.
- `"first"` uses the first period's feature standard deviations in every period.
- `"pooled"` uses standard deviations across all entity-time rows; entities with
  more observations contribute more rows.
- `"period"` estimates standard deviations separately in each period. This
  changes the metric across periods and can absorb real scale changes.

These are substantive metric choices, not uncertainty adjustments. When
standardization is requested, an active feature with zero standard deviation
is rejected. Models must retain at least two positive, numerically identifiable
dimensions. The stored preprocessing records contain the centers and scales.

```{r preprocessing}
standardized <- embed_snapshots(input, features, standardize = "first")
standardized$embedding_metadata$preprocessing$scales[[1]]
```

`feature_weights` supplies the same nonnegative weights in every period. A
weight $w_j$ multiplies the preprocessed feature by $\sqrt{w_j}$; weights are
not normalized. Integer weights therefore reproduce repeated feature columns.
A zero weight removes that feature's geometric contribution.

```{r weights}
weighted <- embed_snapshots(input, features, feature_weights = c(2, 1, 0, 1))
weighted$embedding_metadata$feature_weights
```

Weights provide a building block for a future resampling engine. Setting them
does not run a bootstrap and does not produce standard errors or intervals.
Named variables are not automatically exchangeable sampling units.

## Classical MDS from features or distances

For Euclidean feature distances, PCA and classical MDS yield equivalent
retained geometry up to an orthogonal transformation, apart from numerical
roundoff and nonidentifiable boundary eigenspaces. In the exact rank-2 example,
they preserve the known movement equally well.

```{r cmds-features}
mds_fit <- embed_snapshots(input, features, method = "cmds") |>
  align_snapshots(anchors = letters[1:7])
stopifnot(max(abs(measure_drift(mds_fit)$distance - movement$distance)) < 1e-8)
mds_fit$diagnostics$embedding
```

Alternatively, pass one labeled `dist` object or distance matrix per period.
Matrix row and column names identify entities and must agree in the same
order. Different periods may contain different entities. Use an explicit
ordered period vector, or supply a fully named list whose names give the
ordered period labels.

```{r cmds-distances}
distances <- lapply(1:2, function(period) {
  rows <- input[input$time == period, ]
  x <- as.matrix(rows[features])
  rownames(x) <- rows$entity
  stats::dist(x)
})
mds_from_distances <- embed_snapshots(distances, method = "cmds", periods = 1:2)
mds_from_distances$diagnostics$embedding
```

Classical MDS forms $B=-\tfrac12 J D^{\circ 2}J$, where
$J=I-\mathbf{1}\mathbf{1}^{T}/n$. The two-dimensional coordinates are
$V_2\Lambda_2^{1/2}$ from the two largest positive eigenvalues. Significant
negative eigenvalues indicate that the supplied distances are not Euclidean
at the configured tolerance. The default `negative_eigen = "error"` rejects
that geometry. Explicit `"truncate"` permits an approximation with a warning;
negative eigenvalues are diagnosed and discarded, and no additive distance
correction is applied. This approximation may alter even the supplied pairwise
distances substantially.

Diagnostics report positive and negative inertia, retained inertia fractions,
the relative gap between eigenvalues two and three, and classical-MDS distance
reconstruction error. A tie at the retained/discarded boundary is warned about
because the selected two-dimensional subspace is not unique. A tie wholly
inside the retained two-dimensional space only changes its orientation.

## Reproduce the larger synthetic example

From an installed source checkout, run:

```{bash, eval=FALSE}
Rscript data-raw/simulate-embeddings.R ../embedding-simulation-output
```

The script creates three clusters, 99 observations per period, 111 unique
entities, three periods and eight measured features. One cluster and four
other entities have known movement; some entities disappear and others enter.
Known global rotations, reflection and translation are applied within a fixed
rank-2 feature plane. It saves raw and aligned figures, movement-versus-truth
figures, original measurements, generating transformations, fitted objects and
numerical check results. Separate controls verify zero drift under pure
isometries, PCA/MDS equivalence, distance-input equivalence and integer-weight
column-duplication equivalence.

This idealized example verifies geometry and data flow. It does not establish
robustness under noisy dimension reduction, a correct sampling design, or
confidence-interval coverage. The bootstrap specification is included in
`inst/validation/bootstrap-specification.md` in the source package. A bootstrap
engine, uncertainty intervals, cluster refitting and UMAP adapters are not
implemented.
