---
title: "Getting started with longitudinal coordinate alignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with longitudinal coordinate alignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

Independently calculated embeddings do not necessarily share an orientation or
origin. A map can rotate, reflect, or translate without changing any pairwise
distance. Subtracting such raw coordinates mixes coordinate artifacts with
movement. `driftmapR` fits an explicit common frame before calculating changes.

This prototype accepts supplied two-dimensional coordinates. It estimates
orthogonal Procrustes transformations, measures adjacent-period displacement,
and draws trajectories. It also matches supplied cluster labels with
`match_clusters()`; see the cluster-correspondence vignette. It does not yet fit
PCA or MDS embeddings, quantify bootstrap uncertainty, or support UMAP.

## Construct an analysis

An entity-time pair must identify one row. Supply the ordered period schedule
explicitly when the order or spacing cannot be inferred reliably.

```{r input}
library(driftmapR)

first <- data.frame(
  entity = letters[1:6], time = 1,
  x = c(0, 3, 0, 3, 1, 2), y = c(0, 0, 2, 2, 1, 1)
)
true_second <- transform(first, time = 2)
true_second$x[true_second$entity == "e"] <- 1.4
true_second$y[true_second$entity == "f"] <- 1.2

# Rotate by 90 degrees and translate the second coordinate system.
second <- transform(true_second, x = -y + 7, y = x - 4)
object <- drift_data(rbind(first, second), periods = 1:2)
object
plot_drift_map(object, aligned = FALSE, trajectories = FALSE)
```

The constructor retains row-level metadata such as supplied cluster labels.
Entities may enter or leave the panel. Missing entity identifiers, missing time
values, duplicate entity-time pairs, and invalid coordinates are errors.
A scheduled period with no observations is an error. Omitted periods cannot be
diagnosed from arbitrary labels without a declared schedule; provide `periods`
when that distinction matters.

## Fit the reference frame

For matched entities, let $X$ contain current coordinates and $Y$ contain
reference coordinates. Center both using only the matched fit entities,
$X_c = X - \bar X$ and $Y_c = Y - \bar Y$. Write

$$
X_c^\mathsf{T}Y_c = U \Sigma V^\mathsf{T}.
$$

The orthogonal minimizer is $R = UV^\mathsf{T}$. Reflections are allowed because
independent coordinate systems can differ by a reflection. With `scale = FALSE`,
$s = 1$. With scaling enabled, the least-squares scalar is

$$
s = \frac{\operatorname{tr}(\Sigma)}{\|X_c\|_F^2}.
$$

Using row-vector coordinates, the transformation for every current entity is

$$
X^* = sXR+b, \qquad b=\bar Y-s\bar XR.
$$

New entities receive the same transformation even though they did not help fit
it. The overlap used for fitting must span the coordinate space; insufficient or
degenerate overlap is rejected. Rotation matrices, translations, scale factors,
fit counts, discrepancy, and references are retained in the analysis object.

The default `reference = "previous"` fits each current map to the already
aligned preceding map. This creates one sequential frame. Reference-frame error
can accumulate over a long sequence.

```{r alignment}
# In this synthetic example a:d are known stationary, non-collinear anchors.
aligned <- align_snapshots(object, reference = "previous",
                            scale = FALSE, anchors = letters[1:4])
aligned
aligned$transformations
plot_drift_map(aligned, labels = TRUE)
```

### The reference frame defines what movement means

When `anchors = NULL`, every shared entity contributes to the fitted frame.
That is a least-squares compromise. If many entities move together, the fitted
translation and rotation can absorb part of their genuine movement and induce
apparent movement among otherwise stationary entities. Known stable anchors
preserve movement relative to those anchors, but their stability is an
assumption supplied by the analyst, not established by this function.

Without an external reference, whole-system rigid motion and arbitrary
coordinate changes cannot be distinguished. Optional scale removal similarly
removes whole-system expansion or contraction. Keep `scale = FALSE` when such
changes are meaningful, and examine fit diagnostics before interpreting drift.

## Measure movement programmatically

For shared entities in adjacent scheduled periods,

$$
\Delta_{it}=Z^*_{it}-Z^*_{i,t-1}, \qquad D_{it}=\|\Delta_{it}\|_2.
$$

```{r movement}
movement <- measure_drift(aligned)
movement
stopifnot(max(abs(movement$distance[movement$entity %in% letters[1:4]])) < 1e-9)
stopifnot(abs(movement$distance[movement$entity == "e"] - 0.4) < 1e-9)
stopifnot(abs(movement$distance[movement$entity == "f"] - 0.2) < 1e-9)
```

The output contains entity, source/destination periods, source/destination
coordinates, `dx`, `dy`, and `distance`. A disappearing or entering entity has
no transition at that boundary. A reappearing entity is not connected across an
unobserved intermediate period. The basic plot uses this same adjacency rule.
These quantities are descriptive geometric changes. No inferential detection
threshold or confidence interval is supplied by this prototype.

## Reproduce the larger synthetic example

The source checkout includes `data-raw/simulate-prototype.R`. After installing
the package, run from the checkout:

```{bash, eval=FALSE}
Rscript data-raw/simulate-prototype.R ../simulation-output
```

The script generates three clusters, 104 unique entities, 96 observations in
each of three periods, known coordinate transformations, individual movement,
a moving cluster, and entrants/exits. It compares all-shared alignment against
known-stable-anchor alignment, with and without small coordinate noise. It saves
raw/aligned maps, truth and estimated movement, transformation records, model
objects, numerical summaries, and PNG/PDF figures.

The noiseless anchor case must recover each displacement within an absolute
tolerance of $10^{-9}$ coordinate units. That tolerance is much larger than
machine precision for this small example while far below its imposed movements.
The simulation is one seeded realization at each of two noise levels. It does
not establish interval coverage, false-positive rates, or performance across a
simulation design; those remain later validation work.

## Current limits and next milestones

Cluster-label correspondence is implemented for supplied hard labels. The next
milestone is a defined resampling design for bootstrap movement uncertainty and
cluster stability, supported by original-data adapters. Consensus alignment,
additional plotting functions, a broader simulation study, and applied
vignettes remain separate development work. This prototype is not a CRAN or
R Journal submission.
