---
title: "Paired measurement-unit resampling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paired measurement-unit resampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The first bootstrap engine resamples declared measurement units while keeping
entity identities, the period schedule, and observed availability fixed. It
refits the embedding and alignment in every draw. The outputs are **conditional
resampling summaries**. Calibrated confidence regions, significance tests, and
bootstrap cluster stability are not implemented.

## Declare what was sampled

A measurement unit contains one feature or a disjoint block of features. All
blocks must have equal width. Units must be independent and identically
distributed, or satisfy that assumption within declared strata. Dependence
within a unit, across entities and periods, stays intact. A unit's sampled
multiplicity applies to all of its features in every period.

Named covariates such as age and income are not exchangeable merely because
they are columns in the same table. An unjustified feature bootstrap is a
feature-choice sensitivity experiment. Supply a scientific sampling declaration
with the design; the software cannot verify that declaration.

Here each two-column unit has an independent random loading matrix. The
dependence within its two columns and across periods is intentional.

```{r data}
library(driftmapR)
set.seed(260914L)
entity <- sprintf("E%02d", 1:12)
positions <- cbind(seq(-2, 2, length.out = 12), rnorm(12))
features <- sprintf("f%02d", 1:8)
loadings <- lapply(1:4, function(i) diag(2) + matrix(rnorm(4, sd = 0.18), 2))
panel <- do.call(rbind, lapply(1:3, function(period) {
  latent <- positions
  latent[12, ] <- latent[12, ] + (period - 1) * c(0.35, -0.2)
  measured <- do.call(cbind, lapply(loadings, function(x) latent %*% x))
  colnames(measured) <- features
  data.frame(entity, time = period, measured, check.names = FALSE)
}))
design <- paired_unit_design(
  units = stats::setNames(rep(paste0("unit", 1:4), each = 2), features),
  assumptions = paste(
    "Simulated units have iid random loading matrices; dependence across",
    "columns, entities and periods remains within each unit. Results condition",
    "on these entities and their observed availability."
  )
)
```

Four units and 20 replicates keep this example fast; they are not an adequate
sample-size or simulation-budget recommendation for inference.

## Refit the longitudinal analysis

```{r fit}
fit <- embed_snapshots(panel, features = features) |>
  align_snapshots(anchors = entity[1:11])
result <- bootstrap_drift(fit, design, B = 20L, seed = 914L,
                          keep = "replicates")
result$bootstrap$summary[result$bootstrap$summary$entity == "E12", ]
```

Each replicate first fits its baseline map to the observed baseline with
translation and orthogonal rotation/reflection, **without scaling**. This
expresses displacement vectors in a common observed coordinate frame while
retaining variation in map size. Later maps align to that replicate's previous
or first map, according to the original settings. Stable anchors in this
example are known from the generator; selecting anchors from the data would
require an explicitly conditional or refitted selection procedure.

For `method = "cmds"`, distances are rebuilt from the resampled feature matrix.
A supplied distance matrix alone has no declared underlying measurement units
and is not supported by this design. Preweighted observed feature analyses are
also outside the initial engine's contract.

`preprocess = "refit"` recomputes the original centering/scaling recipe;
`"fixed"` reuses observed calibration. With feature multiplicities alone, the
unchanged feature values and rows make those modes equivalent. The distinction
matters for future resamplers that change measurement values.

## Inspect failures and reproducibility

```{r accounting}
result$bootstrap$attempts
stopifnot(nrow(result$bootstrap$attempts) == 20L,
          all(result$bootstrap$attempts$success))
# This local check also confirms that a bootstrap call preserves caller RNG.
set.seed(19L)
before <- .Random.seed
repeat_result <- bootstrap_drift(fit, design, B = 20L, seed = 914L,
                                 keep = "replicates")
stopifnot(identical(before, .Random.seed),
          identical(result$bootstrap$replicates,
                    repeat_result$bootstrap$replicates))
```

The engine attempts exactly `B` independent L'Ecuyer streams and never retries
failed draws. A failure at any stage invalidates the whole replicate. The
attempt log retains the stage, period, and message. Draw multiplicities and RNG
streams remain available even when only summaries are retained.

The default 95% success gate suppresses estimates if too few draws succeed;
at least two valid draws are always required. Retained successful replicates
can still be inspected. Passing this gate does not remove selection bias from
conditioning on computational success. Check eigengaps, rank/overlap failures,
warning records, and the declared target before interpreting summaries.

## Interpretation stops before inference

Component means, covariance, standard deviations, quantiles, and Monte Carlo
standard errors summarize successful draws. Coordinate and movement tables
carry their denominators. These tables do not contain an automatic significance
decision. In particular, a positive lower percentile for `distance` does not
establish movement: the norm is nondifferentiable at zero and sampled norms are
usually positive.

Calibration requires independent simulated datasets and a declared inferential
target. Coverage, false-positive rates, power, simultaneous inference, angular
summaries, and clustering refits remain future work. See the packaged
`inst/validation/bootstrap-specification.md` for the full design and limitations.
